Finiteness in Cubical Type Theory

### 2020-01-20 paper.tex
more

When we say ``relation'' we are referring to the arrows in
Figure~\ref{finite-classification}.
The proofs, then, amount to a function which inhabits each arrow.
Each unlabelled arrow is a weakening: i.e., every manifest Bishop finite set
is manifest enumerable.
The labelled arrows are strengthening proofs: i.e., every manifest enumerable set
\emph{with decidable equality} is manifest Bishop finite.

We will then characterise each predicate as some form of topos.
Our proofs follow the structure of \cite[Chapters 9, 10]{hottbook} and
\cite{rijkeSetsHomotopyType2015}.
This means we will define first the precategory of sets, and then the category
of sets, and prove the required closures and limits to get us to a topos.
We say ``some form of'' topos here because of course the category of sets in
HoTT do \emph{not}, in fact, form a topos, rather a \(\Pi W\)-pretopos.

\todo[inline]{redo these last two paragraphs}
After the finite predicates, we will briefly look at the infinite countable
types, and classify them in a parallel way to the finite predicates.

All of our work is formalised in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
We will make mention of the few occasions where the formalisation of some proof
is of interest, but the main place where we will discuss the code is in the
final section, where we implement a library for proof search, based on
omniscience and exhaustibility.
We also use this section to demonstrate some \emph{computation} using
univalence: our proof search library can deal seamlessly with functions, testing
them for equality, synthesising them, etc.
\subsection{Related Work}
\subsection{Notation}
\todo[inline]{Basic MLTT Notation: \(\Pi\), \(\Sigma\), etc}
\todo[inline]{Paths, Univalence, isomorphisms}
\todo[inline]{HITs}

\section{Finiteness Predicates}
\subsection{Kuratowski Finiteness}
\begin{rm-definition}[Kuratowski Finite Subset]
  \(\mathcal{K}(A)\) is the type of Kuratowski-finite subsets of \(A\).
  \begin{equation}
    \begin{aligned}
      \mathcal{K}(A) =&
             \; \eta : A \rightarrow \mathcal{K}(A) ; \\
      \vert& \; \emptyset : \mathcal{K}(A) ; \\
      \vert& \; \wc \cup \wc : \mathcal{K}(A) \rightarrow \mathcal{K}(A) \rightarrow \mathcal{K}(A) ; \\
      \vert& \; \text{associative} : \Pi (\mathit{xs}, \mathit{ys}, \mathit{zs}: \mathcal{K}(A)) , (\mathit{xs} \cup \mathit{ys}) \cup \mathit{zs} \equiv \mathit{xs} \cup (\mathit{ys} \cup \mathit{zs}) ; \\
      \vert& \; \text{identity} : \Pi (\mathit{xs}: \mathcal{K}(A)) , \emptyset \cup \mathit{xs} \equiv \mathit{xs} ; \\
      \vert& \; \text{commutative} : \Pi (\mathit{xs}, \mathit{ys}: \mathcal{K}(A)) , \mathit{xs} \cup \mathit{ys} \equiv \mathit{ys} \cup \mathit{xs} ; \\
      \vert& \; \text{idempotent} : \Pi (\mathit{xs}: \mathcal{K}(A)) , \mathit{xs} \cup \mathit{xs} \equiv \mathit{xs} ; \\
      \vert& \; \text{trunc} : \Pi (\mathit{xs}, \mathit{ys}: \mathcal{K}(A)) , \Pi (p, q : \mathit{xs} \equiv \mathit{ys}) , p \equiv q ;
    \end{aligned}
  \end{equation}
\end{rm-definition}
The Kuratowski finite subset is a free join semilattice (or, equivalently, a
free commutative idempotent monoid).
Though it would be possible to use this type as-is, we found it easier to work
with the following type as it has fewer path constructors.
\begin{equation}
  \begin{aligned}
    \mathcal{K}(A) =&
           \; [] : \mathcal{K}(A) ; \\
    \vert& \; \wc \dblcolon \wc : A \rightarrow \mathcal{K}(A) \rightarrow \mathcal{K}(A) ; \\
    \vert& \; \text{com} : \Pi (x, y: A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon y \dblcolon \mathit{xs} \equiv y \dblcolon x \dblcolon \mathit{xs} ; \\
    \vert& \; \text{dup} : \Pi (x : A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon x \dblcolon \mathit{xs} \equiv x \dblcolon \mathit{xs} ; \\
    \vert& \; \text{trunc} : \Pi (\mathit{xs}, \mathit{ys}: \mathcal{K}(A)) , \Pi (p, q : \mathit{xs} \equiv \mathit{ys}) , p \equiv q ;
  \end{aligned}
\end{equation}
What we've done here is basically pulled out the ``monoid'' part of ``free
commutative idempotent monoid'': the first two constructors now construct a
list, which is a representation of the free monoid \emph{without} quotient
types.
This obviates the need for the associative and identity constructors.
We have proven both types equivalent in our formalisation.

@@

\subsection{Manifest Bishop}
\subsection{Manifest Enumerable}
\subsection{Cardinal}
\section{Relations Between Each Finiteness Definition}
\section{Topos}
\section{Countably Infinite Types}
\section{Search}

>>>>

When we say ``relation'' we are referring to the arrows in
Figure~\ref{finite-classification}.
The proofs, then, amount to a function which inhabits each arrow.
Each unlabelled arrow is a weakening: i.e., every manifest Bishop finite set
is manifest enumerable.
The labelled arrows are strengthening proofs: i.e., every manifest enumerable set
\emph{with decidable equality} is manifest Bishop finite.
Our most significant result here is the proof that a cardinal finite set with a
decidable total order is manifestly Bishop finite.

We will then characterise each predicate as some form of topos.
Our proofs follow the structure of \cite[Chapters 9, 10]{hottbook} and
\cite{rijkeSetsHomotopyType2015}.
This means we will define first the precategory of sets, and then the category
of sets, and prove the required closures and limits to get us to a topos.
We say ``some form of'' topos here because of course the category of sets in
HoTT do \emph{not}, in fact, form a topos, rather a \(\Pi W\)-pretopos.
Our main result here is that the category of decidable Kuratowski finite sets
forms a \(\Pi\)-pretopos.

After the finite predicates, we will briefly look at the infinite countable
types, and classify them in a parallel way to the finite predicates.
We here show that countably finite sets form a \(W\)-pretopos.

All of our work is formalised in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
We will make mention of the few occasions where the formalisation of some proof
is of interest, but the main place where we will discuss the code is in the
final section, where we implement a library for proof search, based on
omniscience and exhaustibility.
We also use this section to demonstrate some \emph{computation} using
univalence: our proof search library can deal seamlessly with functions, testing
them for equality, synthesising them, etc.
\subsection{Related Work}
\subsection{Notation}
\todo[inline]{Basic MLTT Notation: \(\Pi\), \(\Sigma\), etc}
\todo[inline]{Paths, Univalence, isomorphisms}
\todo[inline]{HITs}
\section{Finiteness Predicates}
While our primary interest is in decidable Kuratowski finiteness, we found it
useful to examine the other finiteness predicates as well, if only as a way to
construct proofs which can then be transferred to proof on Kuratowski
finiteness.
This, in turn, is why the relation proofs are so important: in HoTT, if two
types are proved equivalent, we get access to the extremely powerful univalence
axiom, which allows us to switch between types freely depending on what is more
convenient.
The \(\Pi\)-topos proof, for instance, is defined on cardinal finiteness, not
decidable Kuratowski finiteness: an equivalence proof between the two types
means we can transfer it over, formally, and with good justification.
Similarly, many of the closure proofs are more naturally defined over the
enumerated finiteness predicates.
\subsection{Kuratowski Finiteness}
The first thing we must define is a representation of subsets.
\begin{rm-definition}[Kuratowski Finite Subset]
  \(\mathcal{K}(A)\) is the type of Kuratowski-finite subsets of \(A\).
  \begin{equation}
    \begin{aligned}
      \mathcal{K}(A) =&
      \; [] : \mathcal{K}(A) ; \\
      \vert& \; \wc \dblcolon \wc : A \rightarrow \mathcal{K}(A) \rightarrow \mathcal{K}(A) ; \\
      \vert& \; \text{com} : \Pi (x, y: A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon y \dblcolon \mathit{xs} \equiv y \dblcolon x \dblcolon \mathit{xs} ; \\
      \vert& \; \text{dup} : \Pi (x : A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon x \dblcolon \mathit{xs} \equiv x \dblcolon \mathit{xs} ; \\
      \vert& \; \text{trunc} : \Pi (\mathit{xs}, \mathit{ys}: \mathcal{K}(A)) , \Pi (p, q : \mathit{xs} \equiv \mathit{ys}) , p \equiv q ;
    \end{aligned}
  \end{equation}
\end{rm-definition}
The Kuratowski finite subset is a free join semilattice (or, equivalently, a
free commutative idempotent monoid).
From a programming perspective, it could well be seen as a finite set (in the
sense of the data structure).
In most languages, it might be represented by a hash table or some balanced
tree: HITs allow us to represent precisely the notion of a finite set, without
putting any constraints on the contents, or choosing any internal data
structure.

Though it would be possible to use this type as-is, we found it easier to work
with the following type as it has fewer path constructors.
What we've done here is basically pulled out the ``monoid'' part of ``free
commutative idempotent monoid'': the first two constructors now construct a
list, which is a representation of the free monoid \emph{without} quotient
types.
This obviates the need for the associative and identity constructors.
We have proven both types equivalent in our formalisation.

@@

\subsection{Manifest Bishop}
\subsection{Manifest Enumerable}
\subsection{Cardinal}
\section{Relations Between Each Finiteness Definition}
\section{Topos}
\begin{itemize}
  \item Complete (pullbacks + terminal object)
\end{itemize}
\section{Countably Infinite Types}
\section{Search}
